Insect-venom elicited anaphylaxis: Data from the European Anaphylaxis Registry

### 2019-08-20 R/insect venom.Rmd
pre revisions from Frontiers

---
title: "Refractory anaphylaxis: Data from the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Macarena Knop^2^
  - Kathrin Scherer Hofmeier^3^
  - Ewa Cichocka-Jarosz^4^
  - Blanca E. García^5^
  - Roland Lang^6^
  - Ioana Maris^7^
  - Jean-Marie Renaudin^8^
  - Margitta Worm^1^ 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: expert-review-of-clinical-immunology.csl # Insert path for the bib-style
#abstract: 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

```{r}
periopAna <- data3$rANA[which(data3$q_152_location=="medical practice, hospital")] %>% 
  as.factor %>%  summary()
nperiopAna <- data3$rANA[which(data3$q_152_location!="medical practice, hospital")] %>% as.factor %>% summary

@@

# Abstract
<!-- 350 Words unstructured -->
Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 µg adrenaline) is a rare and often fatal hypersensitivity reaction. Comprehensive data on its definition, prevalence, and risk factors are missing. Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analyzed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).

The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab[elicitorTab$Elicitor=="drugs","percent"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab[elicitorTab$Elicitor=="drugs","percANA"]`%).  Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees than wasps in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`). The refractory cases occurred mostly in a perioperative setting (`r v$perioperative[3]`% vs. `r v$perioperative[2]`, p < 0.0001). Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases it was applied intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001). Second line treatment options (e.g. vasopression with dopamine, methylene blue, glucagon) were not used at all for the treatment of refractory cases. The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).

Refractory anaphylaxis is associated with drug induced anaphylaxis in particular if given intravenously. Although adrenaline (even i.v.) is frequently applied in these patients, not all patients are responding. Whether a delay in recognition of anaphylaxis is responsible for the refractory case or whether these cases are due to an overflow with mast cell activating substances - requires further studies. Reasons for the low use of second line medication (i.e. methylene blue or dopamine) in refractory cases are unknown, but their use might improve the outcome of severe refractory anaphylaxis cases.

@@

@@

```{r therapyTab}
knitr::kable(therapyTab[,1:4] %>% {data.frame(.[,1],.[,2] %>% as.numeric(),.[,3] %>% as.numeric,.[,4] %>% as.numeric())}, caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. '2nd line' refers to the therapy performed by the professional medical emergency team after the initial rescue procedures. ANA - anaphylaxis, p-value derived from the Fisher's exact test. ",
             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"),
             digits = c(0,2,2,4))
```

@@

>>>>

---
title: "Insect-venom elicited anaphylaxis: Data from the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Macarena Knop^2^
  - Kathrin Scherer Hofmeier^3^
  - Ewa Cichocka-Jarosz^4^
  - Blanca E. García^5^
  - Roland Lang^6^
  - Ioana Maris^7^
  - Jean-Marie Renaudin^8^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::word_document2:
  fig_caption: yes
reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: expert-review-of-clinical-immunology.csl # Insert path for the bib-style
#abstract:
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights.
---

@@

```{r}
periopAna <- data3$rANA[which(data3$q_152_location=="medical practice, hospital")] %>%
  as.factor %>%  summary()
nperiopAna <- data3$rANA[which(data3$q_152_location!="medical practice, hospital")] %>% as.factor %>% summary
```

@@

# Abstract
<!-- 350 Words unstructured -->
Insect-venom elicited anaphylaxis is a common and often severe hypersensitivity reaction. It affects older individuals than other anaphylaxis causes (p = `r t.test(data$d_age[data$d_elicitorInsect=="yes"],data$d_age[data$d_elicitorInsect=="no"])$p.val`). Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified insect-venom elicited anaphylaxis cases (n = `r v$insectElicited`) and analyzed these in comparison to anaphylaxis elicited by other elicitors (n = `r v$otherEli`).

Patients who experienced insect-venom anaphylaxis were older (p < 0.0001), more often had mastocytisis (p < 0.0001) and cardiologic conditions (p < 0.0001) and females more often had concomitant thyroid disease, and less often sufferd from food allergy or atopic dermatitis.

The data show that `r v$insectsExact[7]/sum(v$insectsExact)*100`% of all insect elicited cases were elicited  by yellow jackets, followed by bees (`r v$insectsExact[2]/sum(v$insectsExact)*100`%). The insect venom elicited cases occurred mostly in outdoor places (`r v$locat[4,3]/sum(v$locat[,3])*100`%) patients' homes (`r v$locat[9,3]/sum(v$locat[,3])*100`%) or urban places (`r v$locat[5,3]/sum(v$locat[,3])*100`%). Intramuscular adrenaline (as a first line therapy) was administered significantly less often in insect venom elicited cases (`r therapyTab[1,2]`%, p < 0.0001). The mortality rate in insect anaphylaxis was comparable (`r v$mortality[3]`%) to other cases (`r v$mortality[2]`%, p < `r v$mortality[4]`).

Insect venom anaphylaxis is associated with drug induced anaphylaxis in particular if given intravenously. Although adrenaline (even i.v.) is frequently applied in these patients, not all patients are responding. Whether a delay in recognition of anaphylaxis is responsible for the refractory case or whether these cases are due to an overflow with mast cell activating substances - requires further studies. Reasons for the low use of second line medication (i.e. methylene blue or dopamine) in refractory cases are unknown, but their use might improve the outcome of severe refractory anaphylaxis cases.

@@

## Is elevated serum tryptace aposing risk for severe anaphylaxis in the venom elicited cases more than in other cases?

```{r}
data$d_210_tryptase_cat %>% table(data$d_elicitorInsect)
plot(data$d_severity_rm)

plot(data$d_severity_rm[data$d_210_tryptase_cat=="positive"&data$d_elicitorInsect=="yes"])
plot(data$d_severity_rm[data$d_210_tryptase_cat=="positive"&data$d_elicitorInsect=="no"])
ggplot(data[data$d_210_tryptase_cat=="positive"],)
```

@@

```{r therapyTab}
knitr::kable(therapyTab[,1:4] %>% {data.frame(.[,1],.[,2] %>% as.numeric(),.[,3] %>% as.numeric,.[,4] %>% as.numeric())}, caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. '2nd line' refers to the therapy performed by the professional medical emergency team after the initial rescue procedures. ANA - anaphylaxis, p-value derived from the Fisher's exact test. ",
             col.names = c("Therapy","insect-elicited [%]","other [%]","p value"),
             digits = c(0,2,2,4))
```

@@

```{r age_elicitor_plot, fig.cap="The density of anaphylaxis cases elicited by either insects or other factors according to age"}
age_elicitors_plot
```

```{r tryptase_plot, fig.cap="Tryptase values according to insect elicitors."}
tryptase_plot
```
